Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-03-01 src/tex/ms.tex
plot tweaks and GW170818

\subsection{Hierarchical modeling}

@@

Having established that the collection of detections is globally consistent with $\kappa=0$, here we focus on three events whose $\kappa$ posteriors stand out in Fig.~\ref{fig:posterior_kappa}: GW170818, GW200129, and GW190521.
When considered in isolation, the first of these is the event with the most significant support for $\kappa \neq 0$; the second shows the sign of potential noise anomalies; and the third is representative of a class of events with strongly bimodal $\kappa$ posteriors.

% Case: GW170818
\subsubsection{GW170818}
GW170818 gave the largest tension with $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows the birefringence effect allows a region of $\kappa>0$ for a larger distance and smaller inclination angles than would be allowed in the absence of birefringence.
We can make sense of this result because a smaller inclination angle will make the amplitude ratio between the two circular polarizations smaller.
On the other hand, the birefringence effect will increase the amplitude ratio during propagation.
The larger distance is for compensating the combined effect of a nonzero $\kappa$ and a smaller inclination angle.
This is similar to the degeneracy mentioned in Sec.~\ref{sec:inclination}.
Even though the degeneracy is broken by the frequency dependence to a large extent, it is still possible for getting an extra region with a nonzero $\kappa$.
This is because there is still a significant likelihood in the region of $\kappa$ close to zero.
We can see this effect in other events as well, but not as strongly as in GW170818.
Other parameters such as the spins and polarization angles are notably different from the \ac{GR} case as well.
They may have some slight effect on the degeneracy, but they are not correlated with $\kappa$ significantly.

\begin{figure}
    \script{corner_GW170818.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW170818.pdf}
    \caption{Corner plot of GW170818.}
    \label{fig:corner_GW170818}
\end{figure}

>>>>

\subsection{Hierarchical modeling}
\label{sec:results:hier}

@@

Having established that the collection of detections is globally consistent with $\kappa=0$, here we focus on three events whose $\kappa$ posteriors stand out in Fig.~\ref{fig:posterior_kappa}: GW170818, GW200129, and GW190521.
When considered in isolation, the first of these is the event with the most significant support for nonzero $\kappa$; the second shows the sign of potential noise anomalies; and the third is representative of a class of events with strongly bimodal $\kappa$ posteriors.

% Case: GW170818
\subsubsection{GW170818}
GW170818 produced the posterior in most tension with $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows that this happens because birefringence opens up a region of parameter space with $\kappa>0$ for larger distances and smaller inclination angles than would be allowed in the GR case.
We can make sense of this because an edge-on, non-precessing source will produce linearly-polarized waves, so a smaller inclination will lead the two circular polarizations to have similar amplitudes.
On the other hand, having $\kappa > 0$ will enhance the left-handed modes during propagation.
The two effects can be balanced to yield the correct polarization ratio at the detector, as long as the distance is also enhanced to yield the right amount of birefringence and overall signal power.
This is similar to the degeneracy mentioned in Sec.~\ref{sec:inclination}.
% Even though the degeneracy is broken by the frequency dependence to a large extent, it is still possible for getting an extra region with a nonzero $\kappa$.
% This is because there is still a significant likelihood in the region of $\kappa$ close to zero.
% We can see this effect in other events as well, but not as strongly as in GW170818.

It is not clear what specific feature of the GW170818 data leads to this posterior structure.
However, the reason may be related to the fact that this event provides an uncommonly definite measurement of the polarization angle and spin orientations \cite{Varma:2021csh}.
Indeed, the inferred polarization angle and spins in the birefringent case are are notably different from the \ac{GR} case (Fig. XXX).
If our inference on $\kappa$ is indeed tied to those parameters, we might be susceptible to systematics in the modeling of spin angles in \textsc{IMRPhenomXPHM} and might thus benefit from further analysis with other waveforms like \textsc{NRSur7dq4} \cite{Varma:2018mmi}.
In any case, as we argued in Sec.~\ref{sec:results:hier}, there is no evidence at the moment that this event is inconsistent with the either collection of measurements or with $\kappa = 0$ (Fig.~\ref{fig:reweighted_kappa}).

\begin{figure}
    \script{corner_GW170818.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW170818.pdf}
    \caption{GW170818 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our birefringence analysis (blue), compared to the GR result (orange).}
    \label{fig:corner_GW170818}
\end{figure}
